---
title: "week4-lab"
format: html
---

# Data Viz Hackathon

**The following was obtained from the [README](https://github.com/EDS-240-Data-Viz/fundamental-charts-lab-data/tree/main/Spain_fires) associated with this dataset:**

Background: This dataset is a synthesized master file from the Global Wildfire Information System (GWIS), a joint initiative of the European Commission and the European Space Agency. It combines three critical streams of satellite-derived data for Spain: the number of active fires, the total burnt area, and the resulting atmospheric emissions.

The Data Structure:To make the analysis intuitive, all columns (except for Day) have been prefixed to identify their data source:

-   `nf_` (Number of Fires): Counts of individual thermal anomalies.

-   `ba_` (Burnt Area): Total surface area affected, measured in hectares

-   `em_` (Emissions): Estimated carbon and trace gas release.

More information and metadata can be found [here](https://gwis.jrc.ec.europa.eu/apps/gwis.statistics/seasonaltrend)

Source: Data is sourced from the European Commission’s Joint Research Centre (JRC)

### Read in the data

```{r}
url <- "https://github.com/EDS-240-Data-Viz/fundamental-charts-lab-data/raw/refs/heads/main/Spain_fires/spain_wildfire_joined_2025.csv"
data <- read_csv(url)

# Observe the data's characteristics
glimpse(data)

# 
head(data)
```

### Create a visualization 

```{r, fig.asp =.6}
## Any data wrangling neccessary? 
library(lubridate)
library(ggExtra)
# How have average emissions in 2025 compared to 
data_clean <- data %>% 
  mutate(month = month(dmy(paste0(data$day, "1000")), label = T)) %>% 
  group_by(month) %>% 
  summarize(month_avg = mean(em_average_2003_2025, na.rm = T), 
                                month_avg_2025 = mean(em_year_2025, na.rm = T)) %>% 
  pivot_longer(cols = starts_with("month_avg"), 
               names_to = "period", 
               values_to = "value")
  

data_clean %>% 
  ggplot(aes(x = month, y = value, fill = period)) + 
  geom_col(position = "dodge") +  
  theme_minimal() + labs(x = "Month", y = "Average Emissions", title = ) + theme(legend.position = "bottom")  + scale_fill_manual(values = c(month_avg = "darkred", month_avg_2025 = "darkgreen")) 
  
```

```{r}

```
